---
title: "Ranking speakers by Jensen-Shannon distance and checking Pillai agreement"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ranking speakers by Jensen-Shannon distance and checking Pillai agreement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 5
)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

This vignette walks through the measurement protocol that phontrast
implements in `rank_contrasts()` (since 2.5.0; with the study's calibrated
estimator since 2.5.1). The protocol comes from a simulation study
that scored seven vowel-overlap measures against a known ground truth (Berry,
under review, Sec. VII.A). Its recommendations, in three steps:

1. **Compute Jensen-Shannon distance ($\sqrt{JSD}$) and Pillai on the same
   tokens**, and report the estimated shared probability mass beside them.
2. **Rank speakers by $\sqrt{JSD}$.**
3. **Flag speakers whose Pillai percentile rank differs from their
   $\sqrt{JSD}$ percentile rank by 0.25 of the ordering or more**, inspect
   them, and plot them.

Each step carries conditions: measurements at the $\sqrt{JSD}$ ceiling are set
apart, sample-size floors decide whether a rank or a flag may be read, and a
bandwidth check marks measurements whose rank depends on the smoothing.
`rank_contrasts()` applies and reports all of them.

## The example cohort

`vowel_cohort` is a simulated twelve-speaker cohort built so that each part of
the protocol has something to do: eight speakers with a graded centroid gap,
one speaker whose two vowels share a centroid but differ in shape (a bimodal
"eh"), one speaker with fully separated vowels, and two speakers with fewer
tokens than the others.

```{r cohort}
library(phontrast)

head(vowel_cohort)
table(vowel_cohort$speaker, vowel_cohort$vowel)
```

## Steps 1 to 3 in one call

```{r rank}
ranking <- rank_contrasts(
  data = vowel_cohort,
  features = c("f1", "f2"),
  category_col = "vowel",
  group_col = "speaker"
)
ranking
```

The header states what was computed and under which conditions; the table has
one row per speaker, sorted by $\sqrt{JSD}$. Reading it column by column:

- `sqrt_jsd`, `pillai`, `shared_mass`: step 1. $\sqrt{JSD}$ and shared mass are
  read off one kernel density estimate per speaker, Pillai off the same
  tokens. The kernel estimator is the one the protocol was calibrated on (see
  "The estimator behind the numbers" below).
- `at_ceiling`: `TRUE` where $\sqrt{JSD} \ge 0.99$. spk10's vowels are fully
  separated, so its kernel estimate can no longer grade separation; its shared
  mass and Pillai are still reported, but it leaves both rankings.
- `pr_jsd`, `pr_pillai`, `rank_diff`: step 2, on the percentile-rank scale
  $(\bar r - \tfrac12)/k$ over the $k$ speakers below the ceiling (ties
  averaged).
- `flag`: step 3, `TRUE` where `abs(rank_diff) >= 0.25`. spk09 is the planted
  disagreement: Pillai, a mean-based measure, sees almost no contrast between
  two vowels with the same centroid, while $\sqrt{JSD}$ sees two distributions
  that barely overlap.

```{r flagged}
ranking[ranking$flag %in% TRUE, c("group", "sqrt_jsd", "pillai", "pr_jsd", "pr_pillai", "rank_diff")]
```

## Licensing: when a rank or a flag may be read

The simulation licenses these readings only from certain token counts per
vowel per speaker, and only up to certain dimensionalities. The floors are
applied per speaker from the smaller vowel's count (`n_min`):

```{r floors}
protocol_floors(2)
ranking[, c("group", "n_min", "rank_licensed", "flag_licensed", "rank_basis", "flag")]
```

At two dimensions, ranking by $\sqrt{JSD}$ is licensed from 50 tokens per
vowel and the flag is readable from 100. spk11 (60 tokens) can be ranked by
$\sqrt{JSD}$ but its flag is `NA`: a single speaker's rank difference spans
about 0.28 of the ordering across redraws at 50 tokens, so a 0.25 margin cannot
be read there. spk12 (40 tokens) is below the rank floor, so `rank_basis` says
to order that speaker by Pillai. `rank_diff` is still reported for both; only
the inference is withheld. From eight dimensions the agreement test returns
agreement whatever the data contain, and above eight the study offers ordering
evidence only, so `rank_contrasts()` withholds the flag and the licence there
too.

## The agreement plot

`plot()` on the ranking draws step 3: Pillai percentile rank against
$\sqrt{JSD}$ percentile rank, the identity line, and the inspection band of
$\pm 0.25$ around it. Flagged speakers are coloured and labelled, set-aside
speakers crossed, and speakers whose flag is not readable are hollow.

```{r agreement-plot, eval = has_ggplot2}
plot(ranking)
```

## Inspecting a flagged speaker across bandwidths

Before interpreting a flag, the protocol asks for one more check: recompute
$\sqrt{JSD}$ at half and at twice the diagonal Scott bandwidth (selected once
on the pooled tokens of the pair, as in the study), re-rank, and set the
measurement aside if its rank moves by 0.25 of the ordering or more, or if a
flagged rank difference changes sign. `rank_contrasts()` runs this check by
default (`bw_check = TRUE`):

```{r bw-check}
ranking[, c("group", "sqrt_jsd_half", "sqrt_jsd_double", "bw_shift", "sign_change", "set_aside")]
```

```{r set-aside, include = FALSE}
aside <- ranking$group[ranking$set_aside %in% TRUE]
```

Here the check sets aside `r if (length(aside)) paste(aside, collapse = " and ") else "no speaker"`.
spk09's separation lives in the shape of its bimodal "eh", which twice the
Scott bandwidth smooths away, so its $\sqrt{JSD}$ rank falls when the
bandwidth doubles. The flag stands, but the protocol asks for the
distributions to be inspected before it is interpreted.

`inspect_contrast()` then redraws one speaker with `plot_contrast()`'s layers,
one panel per bandwidth, so the flag can be judged against the distributions
that produced it. The panels are labelled with $\sqrt{JSD}$ and shared mass at
that bandwidth and with the speaker's Pillai, and the subtitle restates the
reported ranks and the outcome of the bandwidth check.

```{r inspect, eval = has_ggplot2, fig.width = 8, fig.height = 4}
inspect_contrast(ranking, "spk09", reverse_x = TRUE, reverse_y = TRUE)
```

The same bandwidth multiplier is available throughout the kernel path as
`bw_scale` (on `jsd_kde_nd()`, `estimate_jsd()`, `phontrast()`,
`plot_contrast()`, and the overlap functions), so any kernel estimate can be
bracketed by hand.

## The estimator behind the numbers

The floors, the margin, the ceiling, and the bandwidth check above were
calibrated on one kernel estimator: phontrast's `method = "legacy"`, which
evaluates both vowels' kernel densities at the pooled tokens and reads
$\sqrt{JSD}$ off the two self-normalized density vectors, with no
leave-one-out correction. Its bandwidth rule, engine, and evaluation subsample
change with dimensionality. `recommended_estimator()` returns the settings,
and `rank_contrasts()` uses them by default:

```{r estimator}
recommended_estimator(2)[c("tier", "method", "bw", "engine", "eval_on", "eval_n", "bracket_bw")]
recommended_estimator(8)[c("tier", "method", "bw", "engine", "eval_on", "eval_n", "bracket_bw")]
```

phontrast's default estimator elsewhere, `method = "mc"`, targets the
continuous divergence and reads higher on overlapping vowels, so it moves
contrasts across the ceiling and through the bandwidth check. Rankings computed
with it are not covered by the study's calibration.

To run the protocol with other settings, pass a modified list:

```{r estimator-override, eval = FALSE}
rank_contrasts(
  vowel_cohort, c("f1", "f2"), "vowel", "speaker",
  estimator = modifyList(recommended_estimator(2), list(bw = "scott.diag", engine = "fast_diag"))
)
```

## Measures and estimators side by side

The study's central distinction is between a *measure* (the quantity) and its
*estimator*. `phontrast()` makes that distinction visible: the closed-form
Gaussian Bhattacharyya columns and the matched-kernel ones estimate the same
quantity under different models, and the whole kernel family (Jensen-Shannon,
overlap, total variation, kernel Bhattacharyya, Hellinger) is scored on one
shared density estimate per comparison.

```{r measures}
one_speaker <- vowel_cohort[vowel_cohort$speaker == "spk05", ]
phontrast(
  one_speaker, c("f1", "f2"), "vowel",
  metrics = c("js_distance", "pillai", "overlap", "tv",
              "bhattacharyya", "bhattacharyya_kde", "euclidean")
)
```

## What to report

For a set of speakers measured on one contrast, report per speaker the three
step-1 quantities (`sqrt_jsd`, `pillai`, `shared_mass`), the token count per
vowel (`n_min`) and the number of features, and then the $\sqrt{JSD}$ ordering
with its licensing. Name the flagged speakers, say whether the flag was readable
at that sample size, and show the inspected distributions for the ones you
discuss. State the estimator settings (`attr(ranking, "protocol")$estimator`)
and the bandwidth-check outcome. `print(ranking)` gives all of this in one
block; the plots above give the picture.
