---
title: "Multidimensional workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multidimensional workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`phonJSD` can estimate category separation in feature spaces with more than two
dimensions, such as MFCC vectors, formants plus duration, or other acoustic
embeddings. The key distinction is that metrics and plots answer different
questions:

- `compare_overlap_metrics()` estimates metrics in the full feature space you
  pass to `features`.
- `plot_category_space()` shows one or two selected dimensions.
- `plot_category_pca()` shows a two-dimensional PCA projection for diagnosing a
  multidimensional space.

```{r}
library(phonJSD)

set.seed(2026)
features <- paste0("feature_", 1:6)
tokens <- data.frame(
  speaker = rep(c("s01", "s02"), each = 80),
  category = rep(rep(c("A", "B"), each = 40), 2),
  matrix(rnorm(160 * length(features)), ncol = length(features))
)
names(tokens)[-(1:2)] <- features
tokens[tokens$category == "B", features[1:3]] <-
  tokens[tokens$category == "B", features[1:3]] + 0.65
```

Compute metrics in all dimensions.

```{r}
metrics <- compare_overlap_metrics(
  data = tokens,
  features = features,
  category_col = "category",
  group_col = "speaker",
  output = "long"
)

metrics[, c("group", "metric", "estimate", "orientation", "separation_value")]
```

Visualize a projection without changing the metric estimand.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot_category_pca(
  data = tokens,
  features = features,
  category_col = "category",
  group_col = "speaker"
)
```

For a specific pair of interpretable dimensions, use `plot_category_space()`.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot_category_space(
  data = tokens,
  features = c("feature_1", "feature_2"),
  category_col = "category",
  group_col = "speaker"
)
```

PCA plots are useful for sanity checks and presentations, but they can hide
separation that lives outside the first two principal components. Use them as a
visual diagnostic; report metric estimates from the intended full feature set.

## PB52 note

The Peterson and Barney 1952 data in `phonTools::pb52` are useful for global
vowel contrasts. For example, a global `I/i` comparison in F1/F2 has many
tokens. Per-speaker `I/i` F1/F2 comparisons are not estimable with KDE-based
metrics because each speaker has only two repetitions per vowel.

```{r, eval = FALSE}
data(pb52, package = "phonTools")
pb_i <- subset(pb52, as.character(vowel) %in% c("I", "i"))

compare_overlap_metrics(
  data = pb_i,
  features = c("f1", "f2"),
  category_col = "vowel"
)
```
