---
title: "Quick start: from vowel table to metrics and plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start: from vowel table to metrics and plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Most analyses should start with a token-level table: one row per observation,
one column for the category contrast, and one or more numeric acoustic
features. `compare_overlap_metrics()` is the preferred entry point because it
returns the main overlap and separation metrics side by side.

```{r}
library(phonJSD)

set.seed(2026)
vowels <- data.frame(
  speaker = rep(c("s01", "s02"), each = 60),
  vowel = rep(rep(c("ih", "eh"), each = 30), 2),
  f1 = c(
    rnorm(30, 500, 55), rnorm(30, 560, 60),
    rnorm(30, 510, 60), rnorm(30, 575, 65)
  ),
  f2 = c(
    rnorm(30, 1980, 150), rnorm(30, 1880, 155),
    rnorm(30, 1960, 160), rnorm(30, 1840, 165)
  )
)
```

The wide output is useful for analysis tables and joining to speaker metadata.

```{r}
metrics_wide <- compare_overlap_metrics(
  data = vowels,
  features = c("f1", "f2"),
  category_col = "vowel",
  group_col = "speaker",
  output = "wide"
)

metrics_wide
```

The long output is easier to rank, filter, and plot. `separation_value` puts
all metrics on a separation-oriented scale: larger values mean greater category
separation, even for overlap metrics such as percent overlap and
Bhattacharyya affinity.

```{r}
metrics_long <- compare_overlap_metrics(
  data = vowels,
  features = c("f1", "f2"),
  category_col = "vowel",
  group_col = "speaker",
  output = "long"
)

metrics_long[, c("group", "metric", "estimate", "orientation", "separation_value")]
```

If `ggplot2` is installed, the same objects can be visualized directly.

```{r, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot_category_space(
  data = vowels,
  features = c("f2", "f1"),
  category_col = "vowel",
  group_col = "speaker",
  reverse_x = TRUE,
  reverse_y = TRUE
)

plot_overlap_metrics(metrics_long)
```

For uncertainty intervals, use `do_boot = TRUE`. Bootstrapping recomputes every
metric on every resample, so use a larger `n_boot` for final analyses than for
interactive examples.

```{r, eval = FALSE}
compare_overlap_metrics(
  data = vowels,
  features = c("f1", "f2"),
  category_col = "vowel",
  group_col = "speaker",
  do_boot = TRUE,
  n_boot = 1000,
  output = "long"
)
```
